Attribute fusion in a latent process model for time series of graphs

### 2011-05-01 ssp2011.tex
Revise proof of scan statistics

@@

  By applying the multivariate normal approximation to the
  multinomial, 

@@

\bibliography{ssp2011}
\end{document}

>>>>

\subsection{Power estimates for $\Psi_{\lambda}(t)$}
The limiting distribution for the scan statistics for unattributed
random graphs was considered in
\cite{rukhin:_limit_distr_graph_scan_statis}. This subsection is
concerned with adpating the results in
\cite{rukhin:_limit_distr_graph_scan_statis} to attributed random
graphs. We state here only the relevant results. The proofs are
delegated to the appendices. \\ \\
The limiting distribution for $\Psi_{\lambda}(t)$ is Gumbel for both
the case where $t < t^{*}$ and for $t = t^{*}$. To aid the exposition,
we first introduce some notations. Let $\lambda^{(2)}$ be the
element-wise square of $\lambda$. Define $C_{00}$, $C_{01}$, $C_{11}$
and $p_{00}$, $p_{01}$, $p_{11}$ to be
  \label{eq:19}
  C_{00} = \tfrac{\langle \lambda^{(2)}, \pi_{00} \rangle}{\langle \lambda,
    \pi_{00}\rangle}; \quad p_{00} = \tfrac{(\langle \lambda, \pi_{00}
    \rangle)^{2}}{\langle \lambda^{(2)},\, \pi_{00} \rangle} \\
C_{11} = \tfrac{\langle \lambda^{(2)}, \pi_{11} \rangle}{\langle \lambda,
    \pi_{11}\rangle}; \quad p_{11} = \tfrac{(\langle \lambda, \pi_{11}
    \rangle)^{2}}{\langle \lambda^{(2)},\, \pi_{11} \rangle} \\
C_{10} = \tfrac{\langle \lambda^{(2)}, \pi_{11} \rangle}{\langle \lambda,
    \pi_{10}\rangle}; \quad p_{10} = \tfrac{(\langle \lambda, \pi_{10}
    \rangle)^{2}}{\langle \lambda^{(2)},\, \pi_{10} \rangle} 
Also let $z_n$ be defined, for any $n \in \mathbb{N}$, by
  \label{eq:20}
   z_n = \sqrt{2
      \log{n}}\Bigl(1 - \frac{\log{\log{n}} + \log{4 \pi}}{4
      \log{n}}\Bigr)
  \label{thm:2}
  Let $\Psi_{\lambda}(t)$ be the scan statistic for $t < t^{*}$. Let
  $N_{0}$, $a_n$, and $b_n$ be defined by
   \mu_0 &= (n-1) \langle 1, \pi_{00} \rangle \\
   \sigma_0 &= \sqrt{(n-1) (\langle 1, \pi_{00} \rangle - \langle
     1, \pi_{00} \rangle^2)} \\ 
    N_{0} &= \mu_0 + z_n \sigma_0 \\
    a_n &= N_{0} \frac{\langle \lambda, \pi_{00} \rangle}{\langle
      1, \pi_{00} \rangle} + C_{00}p_{00}\tbinom{N_{0}}{2} \\ 
   b_n &= (1 - \tfrac{C_{00}p_{00}}{2} + C_{00}p_{00} N_{0}) \frac{\sigma_0}{\sqrt{2 \log
     n}}
Then we have
  \frac{\Psi_{\lambda}(t) - a_{n}}{b_n} \rightsquigarrow
  \mathcal{G}(0, 1)
\end{theorem}%
We now move to the limiting distribution for $\Psi_{\lambda}(t^{*})$. 
Let us define $E$ and $F$ by
 %  B \sim \mathrm{Bin}(m, \langle 1, \pi_{00} \rangle); \quad C \sim \mathrm{Bin}(n-m-1,
 % \langle 1, \pi_{01} \rangle) \\
  E \sim \mathrm{Bin}(m-1, \langle \pi_{11}, 1 \rangle) \\ F \sim \mathrm{Bin}(n-m,
  \langle \pi_{01}, 1 \rangle)
\end{gather*}%
For any random variable $X$, we let $\mu_X$ and $\sigma_X^{2}$ denote
the mean and variance of $X$. If $X$ and $Y$ are independent random
variables, we let $\mu_{X + Y}$ and $\sigma_{X + Y}^2$ denote the mean
and variance of the convolution $X + Y$. \\
\noindent Let $N_{\kappa}$ be the quantity defined by
  \label{eq:18}
  N_{\kappa} = \mu_{E + F} + z_m \sigma_{E + F}
  \label{thm:3}
  Let $a_{n,m}$ and $b_{n,m}$ be given by
    \label{eq:28}
    a_{n,m} &=%  N_{\kappa} \frac{\langle x, \pi_{01} \rangle}{\langle
 %      1, \pi_{01} \rangle} \\
 % & +
    C_{00} p_{00} \Bigl(\tbinom{N_\kappa}{2} - \tbinom{\mu_E}{2} - \mu_E
    \mu_F\Bigr) \\ 
     & + C_{11} p_{11} \tbinom{\mu_E}{2} + C_{01} p_{01} \mu_E \mu_F 
    \end{split} \\
    b_{n,m} = (1 - \tfrac{C_{00} p_{00}}{2} + C_{00} p_{00}
    N_\kappa)\frac{\sigma_{E + F}}{\sqrt{2 \log{n_1}}}
  If $m = \Omega(\sqrt{n \log n})$, then
    \label{eq:29}
    \frac{\Psi_{\lambda}(t^{*}) - a_{n,m}}{b_{n,m}}  \rightsquigarrow \mathcal{G}(0,1)  \end{equation}

@@

  Let $\Gamma^{(k)}(v;t) = |\{w \colon \mathbb{I}\{\phi(wv,t) =
  k\}|$. The vector $\bm{\Gamma}(v;t) = (\Gamma^{(1)}(v;t), \dots,
  \Gamma^{(K)}(v;t))$ for $t < t^{*}$ is distributed as a multinomial
  with $n-1$ trials and probability vector $\pi_{00}$. The folllowing
  statements are made under the assumption that $n \rightarrow \infty$
  for fixed $K$. By the multivariate central limit theorem, we have
    \frac{\bm{\Gamma}(v;t) - (n-1) \pi_{00}}{\sqrt{n-1}}
    \rightsquigarrow \mathcal{N}(0, (\mathrm{diag}(\pi_{00}) -
    \pi_{00} \pi_{00}^{T}))
  Now $d_{\lambda}(v;t) = \langle \lambda, \bm{\Gamma}(v;t) \rangle$
  and we have
    \frac{d_{\lambda}(v;t) - (n-1) \langle \lambda, \pi_{00}
      \rangle}{\sqrt{n-1}} \rightsquigarrow \mathcal{N}(0, \langle \lambda, \eta_{00} \lambda \rangle)
  Therefore, $\tfrac{d_{\lambda}(v;t) - (n-1) \langle \lambda,
    \pi_{00} \rangle}{\sqrt{(n-1) \langle \lambda, \eta_{00} \lambda
      \rangle}} \rightsquigarrow N(0,1)$. We can thus consider the
  degree sequence of $G(t)$ for $t < t^{*}$ as a sequence of {\em
    dependent} Gaussian random variables. By an argument analogous to
  the argument for Erd\"{o}s-Renyi random graphs in \cite[\S
  III.1]{bollobas85:_random_graph} we can show that the dependency
  among the $d_{\lambda}(v,t)$ can be ignored. Let us define
  $\xi_s$ for $s \in \mathbb{R}$ as the number of vertices $v$ such
  that $\tfrac{d_\lambda(v;t) - (n-1) \langle \lambda, \pi_{00}
    \rangle}{\sqrt{(n-1) \langle \lambda, \eta_{00} \lambda \rangle}}
  \geq s$. Note that $\xi_s \leq k$ only if there are at most $k$
  vertices with $\tfrac{d_{\lambda}(v;t) - (n-1) \langle \lambda,
    \pi_{00} \rangle}{\sqrt{(n-1) \langle \lambda, \eta_{00} \lambda
      \rangle}} \geq s$. Thus, $\tfrac{\Delta_{\lambda}(v;t) - (n-1)
    \langle \lambda, \pi_{00} \rangle}{\sqrt{(n-1) \langle \lambda,
      \eta_{00} \lambda \rangle}} < s$ only if $\xi_s = 0$. Because
  the sample maximum of Gaussian random variables converges weakly to
  a Gumbel \cite[\S 2.3]{galambos87:_asymp_theor_extrem_order_statis},
  by taking $n \rightarrow \infty$ followed by $s \rightarrow \infty$,
  we can show that $\tfrac{\Delta_{\lambda}(v;t) - (n-1) \langle \lambda,
  \pi_{00} \rangle}{\sqrt{(n-1) \langle \lambda, \eta_{00} \lambda
    \rangle}} \rightsquigarrow \mathcal{G}(a_n, b_n)$. We thus obtain
Eq.~\eqref{eq:14} for $t < t^{*}$. The case when $t = t^{*}$ can be
derive in a similar manner. We first apply Lemma 3.1 in
\cite{rukhin:_limit_distr_graph_scan_statis} to show that if $m =
\Omega(\sqrt{n \log{n}})$, then $\Delta_{\lambda}(v;t^{*}) \rightsquigarrow
\max_{v \in [m]}{d_{\lambda}(v;t^{*})}$.  
We then show, again by the multivariate
central limit theorem that for $v \in [m]$, $\tfrac{d_{\lambda}(v;t^{*}) -
  \mu_2}{\sigma_2} \rightsquigarrow N(0,1)$. It then follows that
$\max_{v \in [m]} \tfrac{d_{\lambda}(v;t^{*}) -
  \mu_2}{\sigma_2} \rightsquigarrow \mathcal{G}(a_m,b_m)$ and we obtain
Eq.~\eqref{eq:14} for $t = t^{*}$.  

@@

\begin{proof}[Lemma~\ref{lem:2}]
  Let $ X \sim \mathcal{G}(\alpha, \beta)$. We consider the
  normalization $\tfrac{X - \mu}{\sigma}$. We have
    \mathbb{P}[ \frac{X - \mu}{\sigma} \leq z]  &= \mathbb{P}[X \leq z
    \sigma + \mu] 
    = e^{-e^{-(z \sigma + \mu - \alpha)/\beta}} \\
      &= e^{- e^{-(z - (\alpha - \mu)/\sigma)/(\beta/\sigma)}}
  Thus, $\tfrac{X - \mu}{\sigma} \sim \mathcal{G}(\tfrac{\alpha -
    \mu}{\sigma}, \tfrac{\beta}{\sigma})$. Because the sample
  mean and the sample variance are consistent estimators, the claims
  follow after an application of Slutsky's theorem.
\begin{proof}[Theorem~\ref{thm:2}]
Let $\phi_{\lambda}(v;t) = \psi_{\lambda}(v;t) - d_{\lambda}(v;t)$ be
the (fused) locality statistics for
vertex $v$ at time $t$ not including the (fused) degree of $v$, i.e.,
  \label{eq:11}
  \phi_{\lambda}(v;t) = \sum_{k=1}^{K} \lambda_k \sum_{\substack{uw
      \in N(v) \\ u,w \not = v}}
  \mathbb{I}\{\phi(uw,t) = k\}.
The following statements are made conditional on $|N(v)| = l$. First
of all, we have
  \phi_{\lambda}(v;t) = \sum_{k=1}^{K}{\lambda_k z_k}
where the $(z_1, \dots, z_K)$ are distributed as
  (z_1,z_2,\dots,z_K) \sim \textrm{multinomial}\Bigl(
  \tbinom{l}{2}, \pi_{00}\Bigr) 
By the multivariate CLT, we have
  \frac{\phi_\lambda(v;t) - \tbinom{l}{2} \langle \lambda, \pi_{00}
    \rangle}{\sqrt{\tbinom{l}{2} \langle \lambda, \eta_{00} \lambda \rangle}}
 \rightsquigarrow \mathcal{N}(0,1) 
Let $Y_{l} = C_{00} \mathrm{Bin}(\tbinom{l}{2}, p_{00})$. Then $\mathbb{E}[Y_l] = \tbinom{l}{2} \langle \lambda, \pi_{00} \rangle$
and $\mathrm{Var}[Y_l] = \tbinom{l}{2} \langle \lambda,
\eta_{00} \lambda \rangle$ and by the CLT, we have
  \label{eq:12}
  \frac{\psi_{\lambda}(v;t) - \tbinom{l}{2} \langle \lambda, \pi_{00}
    \rangle}{\sqrt{\tbinom{l}{2} \langle \lambda, \eta_{00} \lambda
      \rangle}} \rightsquigarrow \frac{Y_l - \tbinom{l}{2} \langle \lambda, \pi_{00} \rangle}{\sqrt{\tbinom{l}{2}
    \langle \lambda, \eta_{00} \lambda \rangle}}
Eq.~\eqref{eq:12} states that the locality statistics for our
attributed random graphs model with $t < t^{*}$
can be approximated by the locality statistics for an Erd\"{o}s-Renyi
graph with edge probability $p_{00}$. Theorem~\ref{thm:2} then follows
from Theorem~1.1 in \cite{rukhin:_limit_distr_graph_scan_statis}.
\begin{proof}[Theorem~\ref{thm:3}]
  Let $\phi_{\lambda}(v;t^{*}) = \psi_{\lambda}(v;t^{*}) -
  d_{\lambda}(v;t^{*})$. Let $d_{\zeta}(v;t)$ be
the number of neighbors of $v$ that lies in $[m]$ and
$d_{\xi}(v;t)$ be the number of neighbors of $v$ that lies in $[n]
\setminus [m]$. The following statements are made, conditional on $d_{\zeta}(v;t^{*}) =
l_{\zeta}$ and $d_{\xi}(v;t^{*}) = l_{\xi}$. First of all, we have
  \phi_{\lambda}(v;t^{*}) = \sum_{k=1}^{K} \lambda_k ( y^{(\zeta)}_k +
  y^{(\xi)}_k + y^{(\omega)}_k)
where $(y^{(\zeta)}_1, \dots, y^{(\zeta)}_K)$, $(y^{(\xi)}_1,\dots,
 y^{(\xi)}_K)$,  $(y^{(\omega)}_1, \dots, y^{(\omega)}_K)$ are
 distributed as
(y^{(\zeta)}_1,\dots,y^{(\zeta)}_K) \sim \textrm{multinomial}\Bigl(
\tbinom{l_\zeta}{2}, \pi_{11}\Bigr) \\ 
(y^{(\xi)}_1,\dots,y^{(\xi)}_K) \sim \textrm{multinomial}\Bigl(
\tbinom{l_\xi}{2}, \pi_{00}\Bigr) \\
(y^{(\omega)}_1,\dots,y^{(\omega)}_m) \sim \textrm{multinomial}\Bigl(
l_\zeta l_\xi, \pi_{10}\Bigr)
Let $\rho$ and $\varsigma$ be defined as
  \rho = \langle \lambda, \tbinom{l_{\zeta}}{2} \pi_{11} +
  \tbinom{l_{\xi}}{2} \pi_{00} + l_{\zeta} l_{\xi} \pi_{10} \rangle \\
  \varsigma = \langle \lambda, \Bigl(\tbinom{l_{\zeta}}{2} \eta_{11} +
  \tbinom{l_{\xi}}{2} \eta_{00} + l_{\zeta} l_{\xi} \eta_{10}\Bigr) \lambda
  \rangle
By the multivariate CLT, we have
  \label{eq:16}
  \frac{\phi_{\lambda}(v;t^{*}) - \rho}{\varsigma} \rightsquigarrow
  \mathcal{N}(0,1)
Let $Y_{\zeta} \sim C_{11} \mathrm{Bin}\Bigl(\tbinom{l_{\zeta}}{2},
p_{11}\Bigr)$, $Y_{\xi} \sim C_{00} \mathrm{Bin}\Bigl(\tbinom{l_\xi}{2},
p_{00}\Bigr)$ and $Y_{\omega} \sim C_{10} \mathrm{Bin}\Bigl( l_{\zeta}
l_{\xi}, p_{10} \Bigr)$. By the CLT, we have
  \label{eq:21}
 \frac{\phi_{\lambda}(v;t^{*}) - \rho}{\varsigma} \rightsquigarrow
 \frac{Y_{\zeta} + Y_{\xi} + Y_{\omega} - \rho}{\varsigma}
Eq.~\eqref{eq:21} states that the locality statistics for our
attributed random graphs model time $t = t^{*}$ can be approximated by
the locality statistics for an unattributed kidney and egg model.
\bibliography{ssp2011}
\end{document}
